---
title: "flat_full.Rmd for working package"
output: html_document
editor_options: 
  chunk_output_type: console
---

<!-- Run this 'development' chunk -->
<!-- Store every call to library() that you need to explore your functions -->

```{r development, include=FALSE}
library(testthat)
```

<!--
 You need to run the 'description' chunk in the '0-dev_history.Rmd' file before continuing your code there.

If it is the first time you use {fusen}, after 'description', you can directly run the last chunk of the present file with inflate() inside.
--> 

```{r development-load}
# Load already included functions if relevant
pkgload::load_all(export_all = FALSE)
```

<!-- 
 Store your dataset in a directory named "inst/" at the root of your project.
 Use it for your tests in this Rmd thanks to `pkgload::load_all()` to make it available
and `system.file()` to read it in your examples.

- There already is a dataset in the "inst/" directory to be used in the examples below
-->

# Importing generated reports from the Bruker MALDI Biotyper device

## Importing a single report

<!--
Create a chunk for the core of the function

- The chunk needs to be named `function` at least
- It contains the code of a documented function
- The chunk can also be named `function-my_median` to make it easily
findable in your Rmd
- Let the `@examples` part empty, and use the next `examples` chunk instead to present reproducible examples

After inflating the template

-  This function code will automatically be added in a new file in the "R/" directory
-->

```{r function_read_biotyper_report}
#' Importing Bruker MALDI Biotyper CSV report
#'
#'
#' The headerless table exported by the Compass software in the Bruker MALDI
#' Biotyper device is separated by semi-colons and has empty columns which prevent
#' an easy import in R. This function reads the report correctly as a tibble.
#' 
#' @details
#' The headerless table contains identification information for each target processed by
#' the Biotyper device and once processed by the `read_biotyper_report`,
#' the following seven columns are available in the tibble, _when using the `best_hits = TRUE` option:
#' * `spot`: an integer indicating the spot number of the MALDI target (i.e., plate)
#' * `sample_name`: the character string provided during the preparation of the MALDI target (i.e., plate)
#' * `bruker_quality`: a character encoding the quality of the identification with potentially multiple "+" symbol or only one "-"
#' * `bruker_species`: the species name associated with the MALDI spectrum analyzed.
#' * `bruker_taxid`: the NCBI Taxonomy Identifier of the species name in the column species
#' * `bruker_hash`: a hash from an undocumented checksum function probably to encode the database entry.
#' * `bruker_log`: the log-score of the identification.
#' 
#' When all hits are returned (with `best_hits = FALSE`), the two columns `spot` and `sample_name`
#' remains unchanged, but the five columns prefixed by `bruker_` contains the hit number:
#'
#' * `bruker_01_quality`
#' * `bruker_01_species`
#' * `bruker_01_taxid`
#' * `bruker_01_hash`
#' * `bruker_01_log`
#' * `bruker_02_quality`
#' * ...
#' * `bruker_10_species`
#' * `bruker_10_taxid`
#' * `bruker_10_hash`
#' * `bruker_10_log`
#'
#' @param path Path to the semi-colon separated table
#' @param best_hits A logical indicating whether to return only the best hit
#'
#' @return
#' A tibble of 7 columns (`best_hits = TRUE`) or 52 columns (`best_hits = FALSE`). See Details for the description of the columns.
#' @export
#'
#' @examples
read_biotyper_report <- function(path, best_hits = TRUE){
  # Prepare the columns names, because 10 hits are reported by default
  prep_names <- tidyr::expand_grid(
    "prefix" = "bruker",
    "iteration" = sprintf("%02d", 1:10), # Because 10 hits per spot with each 5 columns
    "variables" = c("quality", "species", "taxid", "hash", "log")
  ) %>% dplyr::mutate(
    "type" = dplyr::if_else( .data$variables == "log", "d", "c"),
    "col_names" = paste(.data$prefix, .data$iteration, .data$variables, sep = "_")
  )
  
  # Read in the report, usually warnings about problems and
  #  inconsistent number of columns are triggered
  breport <- utils::read.delim(
    path, 
    col.names = c("spot", "sample_name", prep_names$col_names),
    sep = ";", header = FALSE,
    na = c("NA", "E1", "E2", "") # Added E1 identification in taxid as NA
  )
  no_peak_lgl <- breport$bruker_01_species == "no peaks found"
  
  # Remove the spot for which no peaks were detected, and warn the user
  breport <- tibble::as_tibble(breport) %>%
    dplyr::filter(.data$bruker_01_species != "no peaks found")
  if(sum(no_peak_lgl) > 0 ){
    warning(
      "Remove ", sum(no_peak_lgl), " row(s) out of ", length(no_peak_lgl),
      " due to no peaks found")    
  }

  if(best_hits){
    breport %>%
      # Replaced .data$spot by "spot"
      # src: https://tidyselect.r-lib.org/news/index.html#lifecycle-changes-1-2-0
      dplyr::select("spot", "sample_name", tidyselect::starts_with("bruker_01")) %>%
      dplyr::rename_with(~ gsub("_01", "", .x)) %>%
      return()
  } else{
    return(breport)
  }
}
```

<!--
Create a chunk with an example of use for your function

- The chunk needs to be named `examples` at least
- It contains working examples of your function
- The chunk is better be named `examples-my_median` to be handled
correctly when inflated as a vignette

After inflating the template

-  This example will automatically be added in the '@examples' part of our function above in the "R/" directory
- This example will automatically be added in the vignette created from this Rmd template
-->

```{r examples_read_biotyper_report}
# Get a example Bruker report
biotyper <- system.file("biotyper.csv", package = "maldipickr")
# Import the report as a tibble
report_tibble <- read_biotyper_report(biotyper)
# Display the tibble
report_tibble
```

<!--
Create a chunk with a test of use for your function

- The chunk needs to be named `tests` at least
- It contains working tests of your function
- The chunk is better be named `tests-my_median` to be handled
correctly when inflated as a vignette

After inflating the template

-  This test code will automatically be added in the "tests/testthat/" directory
-->

```{r tests_read_biotyper_report}
# test_that("my_median works properly and show error if needed", {
#   expect_true(my_median(1:12) == 6.5)
#   expect_error(my_median("text"))
# })
# Test with a correct and empty datasets in "inst/"
biotyper <- system.file("biotyper.csv", package = "maldipickr")
biotyper_empty <- system.file("biotyper_empty.csv", package = "maldipickr")

# Apply test on my function
test_that("read_biotyper_report works properly with correct dataset and best hits", {
  expect_equal(
    nrow(read_biotyper_report(biotyper)), 3
  )
  expect_equal(
    ncol(read_biotyper_report(biotyper)), 7
  )
  expect_equal(
    dplyr::pull(read_biotyper_report(biotyper), bruker_species),
    c("not reliable identification", "Escherichia coli", "Kosakonia cowanii")
  )
})
test_that("read_biotyper_report works properly with correct dataset and all hits", {
  expect_equal(
    nrow(read_biotyper_report(biotyper, best_hits = FALSE)), 3
  )
  expect_equal(
    ncol(read_biotyper_report(biotyper, best_hits = FALSE)), 52
  )
})
test_that("read_biotyper_report is empty when no peaks are found", {
  expect_equal(
    nrow(read_biotyper_report(biotyper_empty)), 0
  )
  expect_warning(
    read_biotyper_report(biotyper_empty),"Remove"
    )
})
```

## Importing multiple reports

```{r function-read_many_biotyper_reports}
#' Importing a list of Bruker MALDI Biotyper CSV reports
#'
#' @param path_to_reports A vector of paths to the csv files to be imported by [read_biotyper_report()].
#' @param report_ids A vector of character names for each of the reports.
#' @param best_hits A logical indicating whether to return only the best hit in the [read_biotyper_report()] function.
#' @param ... Name-value pairs to be passed on to [dplyr::mutate()]
#'
#' @return A tibble just like the [read_biotyper_report()] function except for an additional column `name` with the `report_ids` used as a prefix of the `spot` name.
#' 
#' @seealso [read_biotyper_report()]
#' 
#' @note The report identifiers are sanitised to convert all dashes (`-`) as underscores (`_`). 
#' 
#' @export
#'
#' @examples
read_many_biotyper_reports <- function(path_to_reports, report_ids, best_hits = TRUE, ...){
  # Import the Bruker Biotyper reports as a named list
  breports <- lapply(
    path_to_reports,
    read_biotyper_report,
    best_hits
  )
  names(breports) <- report_ids
  # Conversion of a named list of dataframe to the dataframe with the name as
  #  a column is now super easy with enframe()
  tibble::enframe(breports) %>% tidyr::unnest("value") %>%
    dplyr::mutate(
      "name" = paste(gsub("-", "_", .data$name), .data$spot, sep = "_"),
      "bruker_species" = if_else(
        .data$bruker_species == "not reliable identification",
        paste(.data$bruker_species, .data$name), # make the unmatched species unique for later counts
        .data$bruker_species
      ),
      ...
    )  %>% return()
}
```
  
```{r example-read_many_biotyper_reports}
# List of Bruker MALDI Biotyper reports
reports_paths <- system.file(
  c("biotyper.csv", "biotyper.csv", "biotyper.csv"),
  package = "maldipickr")
read_many_biotyper_reports(
  reports_paths,
  report_ids = c("first","second","third"),
  growth_temperature = 37.0 # Additional metadata passed to dplyr::mutate                 
)
```
  
```{r tests-read_many_biotyper_reports}
dummy_multiple_paths <- system.file(
  c("biotyper.csv","biotyper_empty.csv", "biotyper.csv"),
  package = "maldipickr")
test_that("read_many_biotyper_reports works", {
  expect_equal(
    nrow(
      read_many_biotyper_reports(
        dummy_multiple_paths,
        report_ids = c("first","second","third")
      )
    ), 6
  )
})
test_that("read_many_biotyper_reports works with arguments passed to mutate", {
  expect_equal(
    read_many_biotyper_reports(
      dummy_multiple_paths,
      report_ids = c("first","second","third"),
      growth_temperature = 37.0
    ) %>%
      dplyr::pull(growth_temperature) %>% unique(), 37 
  )
})
```

# Importing spectra from the Bruker MALDI Biotyper device

## Importing multiple spectra from a directory hierarchy

```{r function-import_biotyper_spectra}
#' Importing spectra from the Bruker MALDI Biotyper device
#' 
#' This function is a wrapper around the [MALDIquantForeign::importBrukerFlex()] to read both `acqus` and `acqu` MALDI files.
#' 
#' 
#' @details When using [MALDIquantForeign::importBrukerFlex()] on `acqus` files (instead of the native `acqu` files), the function will fail with the following error message:
#' 
#' ```
#' Error in .readAcquFile(fidFile = fidFile, verbose = verbose) :
#' File ‘/data/maldi_dir/targetA/0_D10/1/1SLin/acqu’ doesn't exists!
#' ```
#' 
#' But it turns out that `acqu` and `acqus` files [are the same]([https://github.com/sgibb/readBrukerFlexData/wiki/acqu-file), so the function here create `acqu` symbolic links that point to `acqus` files.
#' 
#' @param biotyper_directory A path to the folder tree with the spectra to be imported.
#' @param remove_calibration A vector of characters used as regex to indicate which (calibration) spectra are going to be removed.
#'
#' @return A list of [MALDIquant::MassSpectrum] objects
#' 
#' @export
import_biotyper_spectra <- function(biotyper_directory, remove_calibration = c("BTS", "Autocalibration")){
  # List all the acqus files
  acqus_files <- list.files(
    biotyper_directory, pattern = "acqus$",
    recursive = TRUE, full.names = TRUE)
  # Prepare the absolute filename for acqu
  acqu_files <- gsub("acqus$","acqu", acqus_files)
  # Acqu files as symbolic links if they do not exist
  do_symbolic_links <- !file.exists(acqu_files)
  if(any(do_symbolic_links)){
    links_status <- file.symlink(
      from = acqus_files[do_symbolic_links],
      to = acqu_files[do_symbolic_links]
    )
    if(any(!links_status)){
      stop("Symbolic links could not be created for the files:",
           acqus_files[!links_status]
      )
    }
  }
  biotyper_list <- MALDIquantForeign::importBrukerFlex(biotyper_directory)
  # Catch the arguments of which spectrum to remove
  remove_calibration <- match.arg(remove_calibration, several.ok = T)
  if(length(remove_calibration) != 0){
    # Regex to remove the chosen spectrum
    to_catch <- c("BTS" = "BTS|BTS_Validation",
                  "Autocalibration" = "Autocalibration")
    regex_to_catch <- match.arg(remove_calibration, to_catch, several.ok = T) %>%
      paste(collapse = "|")
    to_remove <- sapply(biotyper_list, function(x) grepl(regex_to_catch, MALDIquant::metaData(x)$file))
    biotyper_list <- biotyper_list[!to_remove]
  }
  return(biotyper_list)
}
```
  
```{r example-import_biotyper_spectra}
directory_for_biotyper_spectra <- system.file("toy-species-spectra", package = "maldipickr")
spectra_list <- import_biotyper_spectra(directory_for_biotyper_spectra)
spectra_list
```
  
```{r tests-import_biotyper_spectra}
directory_for_biotyper_spectra <- system.file("toy-species-spectra", package = "maldipickr")
test_that("import_biotyper_spectra works", {
  expect_length(
    import_biotyper_spectra(directory_for_biotyper_spectra),
    6
  ) 
})
```

## Evaluate the quality of the spectra


```{r function-check_spectra}
#' Evaluate the spectra regularities
#'
#' Assess whether all the spectra in the list are not empty, of the same length and correspond to profile data.
#'
#' @param spectra_list A list of [MALDIquant::MassSpectrum] objects
#'
#' @return A list of logical vectors of length `spectra_list` indicating if the spectra are empty (`is_empty`), of an odd length (`is_outlier_length`) or not a profile spectra (`is_not_regular`).
#'
#' @export
check_spectra <- function(spectra_list) {
  # Checking spectra are not empty
  empty_spectra <- sapply(spectra_list, MALDIquant::isEmpty)
  # Checking spectra are the same length
  ## Getting the most common length
  common_length <- sapply(spectra_list, length) %>%
    # gives length as character/names and occurrence as value
    table() %>%
    which.max() %>%
    # extract length
    names() %>%
    strtoi()
  length_spectra <- sapply(spectra_list, length) == common_length
  # Checking spectra are profile data
  regular_spectra <- sapply(spectra_list, MALDIquant::isRegular)
  # Summarise the checks
  checking_list <- list(
    is_empty = empty_spectra,
    is_outlier_length = !length_spectra,
    is_not_regular = !regular_spectra
  )
  if (any(Reduce(any, checking_list))) {
    # List all the spectra that should be removed
    to_remove <- lapply(checking_list, which) %>%
      unlist() %>%
      unique() %>%
      paste(collapse = ",")
    warning(
  "Some spectra (", to_remove, ") are incorrect (empty, outlier length or irregular).\n",
  "They can be removed as follow:\n",
  " spectra_list <- spectra_list[-c(", to_remove, ")]"
)
  }
  return(checking_list)
}
```
  
```{r example-check_spectra}
directory_for_biotyper_spectra <- system.file("toy-species-spectra", package = "maldipickr")
spectra_list <- import_biotyper_spectra(directory_for_biotyper_spectra)
check_spectra(spectra_list)
```
  
```{r tests-check_spectra}
test_that("check_spectra works", {
  expect_type(
    check_spectra(
      c(MALDIquant::createMassSpectrum(1:3, 1:3), MALDIquant::createMassSpectrum(11:13, 11:13))
    ), "list"
  )
})
test_that("check_spectra warns with empty spectra", {
  expect_warning(
    check_spectra(c(MALDIquant::createMassSpectrum(mass = 0, intensity = 0))),
    "Some spectra"
  )
})
```
  

```{r development-inflate, eval=FALSE}
# Run but keep eval=FALSE to avoid infinite loop
# Execute in the console directly
fusen::inflate(flat_file = "dev/flat_full.Rmd", vignette_name = "Import data from Bruker MALDI Biotyper")
```
