---
title: "01-import-data-from-bruker-maldi-biotyper"
output: html_document
editor_options: 
  chunk_output_type: console
---

<!-- Run this 'development' chunk -->
<!-- Store every call to library() that you need to explore your functions -->

```{r development, include=FALSE}
library(testthat)
```

<!--
 You need to run the 'description' chunk in the '0-dev_history.Rmd' file before continuing your code there.

If it is the first time you use {fusen}, after 'description', you can directly run the last chunk of the present file with inflate() inside.
--> 

```{r development-load}
# Load already included functions if relevant
pkgload::load_all(export_all = FALSE)
```

<!-- 
 Store your dataset in a directory named "inst/" at the root of your project.
 Use it for your tests in this Rmd thanks to `pkgload::load_all()` to make it available
and `system.file()` to read it in your examples.

- There already is a dataset in the "inst/" directory to be used in the examples below
-->

The matrix-assisted laser desorption/ionization-time-of-flight (MALDI-TOF) technology is coupled with mass spectrometry in the Bruker MALDI Biotyper device in order to identify microorganisms.
The device generates two types of data:

1. A report of the identification using its proprietary database of mass spectrum projections (MSPs).
2. The raw mass spectrometry data.

The following vignette describe how to streamline the import of these two types of data into R using the [`maldipickr`](https://github.com/cpauvert/maldipickr) package

# Importing generated reports from the Bruker MALDI Biotyper device

## Importing a single report

The Bruker MALDI Biotyper generates a report via the Compass software summarizing the identification of the microorganisms using its internal database.
While the file is separated by semi-colons, it contains no headers.
The report has many columns in a long format to describe the ten hits when identification is feasible, or only a few when no identification was possible.
All-in-all, this makes the table import into R and its manipulation relatively painful.

Below is an example of import of a single Bruker MALDI Biotyper report into a [`tibble`](https://tibble.tidyverse.org) for use with the [`tidyverse`](https://tidyverse.tidyverse.org/) suite. By default, only the best hit of each colony is reported.

<!--
Create a chunk for the core of the function

- The chunk needs to be named `function` at least
- It contains the code of a documented function
- The chunk can also be named `function-my_median` to make it easily
findable in your Rmd
- Let the `@examples` part empty, and use the next `examples` chunk instead to present reproducible examples

After inflating the template

-  This function code will automatically be added in a new file in the "R/" directory
-->

```{r function_read_biotyper_report}
#' Importing Bruker MALDI Biotyper CSV report
#'
#'
#' The headerless table exported by the Compass software in the Bruker MALDI
#' Biotyper device is separated by semi-colons and has empty columns which prevent
#' an easy import in R. This function reads the report correctly as a tibble.
#'
#' @details
#' The headerless table contains identification information for each target processed by
#' the Biotyper device and once processed by the `read_biotyper_report`,
#' the following seven columns are available in the tibble, _when using the `best_hits = TRUE` option:
#' * `spot`: an integer indicating the spot number of the MALDI target (i.e., plate)
#' * `sample_name`: the character string provided during the preparation of the MALDI target (i.e., plate)
#' * `bruker_quality`: a character encoding the quality of the identification with potentially multiple "+" symbol or only one "-"
#' * `bruker_species`: the species name associated with the MALDI spectrum analyzed.
#' * `bruker_taxid`: the NCBI Taxonomy Identifier of the species name in the column species
#' * `bruker_hash`: a hash from an undocumented checksum function probably to encode the database entry.
#' * `bruker_log`: the log-score of the identification.
#'
#' When all hits are returned (with `best_hits = FALSE`), the two columns `spot` and `sample_name`
#' remains unchanged, but the five columns prefixed by `bruker_` contains the hit number:
#'
#' * `bruker_01_quality`
#' * `bruker_01_species`
#' * `bruker_01_taxid`
#' * `bruker_01_hash`
#' * `bruker_01_log`
#' * `bruker_02_quality`
#' * ...
#' * `bruker_10_species`
#' * `bruker_10_taxid`
#' * `bruker_10_hash`
#' * `bruker_10_log`
#'
#' @param path Path to the semi-colon separated table
#' @param best_hits A logical indicating whether to return only the best hit
#'
#' @return
#' A tibble of 7 columns (`best_hits = TRUE`) or 52 columns (`best_hits = FALSE`). See Details for the description of the columns.
#' @export
#'
#' @examples
read_biotyper_report <- function(path, best_hits = TRUE) {
  # Prepare the columns names, because 10 hits are reported by default
  prep_names <- tidyr::expand_grid(
    "prefix" = "bruker",
    "iteration" = sprintf("%02d", 1:10), # Because 10 hits per spot with each 5 columns
    "variables" = c("quality", "species", "taxid", "hash", "log")
  ) %>% dplyr::mutate(
    "type" = dplyr::if_else(.data$variables == "log", "d", "c"),
    "col_names" = paste(.data$prefix, .data$iteration, .data$variables, sep = "_")
  )

  # Read in the report, usually warnings about problems and
  #  inconsistent number of columns are triggered
  breport <- utils::read.delim(
    path,
    col.names = c("spot", "sample_name", prep_names$col_names),
    sep = ";", header = FALSE,
    na = c("NA", "E1", "E2", "") # Added E1 identification in taxid as NA
  )
  no_peak_lgl <- breport$bruker_01_species == "no peaks found"

  # Remove the spot for which no peaks were detected, and warn the user
  breport <- tibble::as_tibble(breport) %>%
    dplyr::filter(.data$bruker_01_species != "no peaks found")
  if (sum(no_peak_lgl) > 0) {
    warning(
      "Remove ", sum(no_peak_lgl), " row(s) out of ", length(no_peak_lgl),
      " due to no peaks found"
    )
  }

  if (best_hits) {
    breport %>%
      # Replaced .data$spot by "spot"
      # src: https://tidyselect.r-lib.org/news/index.html#lifecycle-changes-1-2-0
      dplyr::select("spot", "sample_name", tidyselect::starts_with("bruker_01")) %>%
      dplyr::rename_with(~ gsub("_01", "", .x)) %>%
      return()
  } else {
    return(breport)
  }
}
```

<!--
Create a chunk with an example of use for your function

- The chunk needs to be named `examples` at least
- It contains working examples of your function
- The chunk is better be named `examples-my_median` to be handled
correctly when inflated as a vignette

After inflating the template

-  This example will automatically be added in the '@examples' part of our function above in the "R/" directory
- This example will automatically be added in the vignette created from this Rmd template
-->

```{r examples_read_biotyper_report}
# Get a example Bruker report
biotyper <- system.file("biotyper.csv", package = "maldipickr")
# Import the report as a tibble
report_tibble <- read_biotyper_report(biotyper)
# Display the tibble
report_tibble
```

<!--
Create a chunk with a test of use for your function

- The chunk needs to be named `tests` at least
- It contains working tests of your function
- The chunk is better be named `tests-my_median` to be handled
correctly when inflated as a vignette

After inflating the template

-  This test code will automatically be added in the "tests/testthat/" directory
-->

```{r tests_read_biotyper_report}
# test_that("my_median works properly and show error if needed", {
#   expect_true(my_median(1:12) == 6.5)
#   expect_error(my_median("text"))
# })
# Test with a correct and empty datasets in "inst/"
biotyper <- system.file("biotyper.csv", package = "maldipickr")
biotyper_empty <- system.file("biotyper_empty.csv", package = "maldipickr")

# Apply test on my function
test_that("read_biotyper_report works properly with correct dataset and best hits", {
  expect_equal(
    nrow(read_biotyper_report(biotyper)), 3
  )
  expect_equal(
    ncol(read_biotyper_report(biotyper)), 7
  )
  expect_equal(
    dplyr::pull(read_biotyper_report(biotyper), bruker_species),
    c("not reliable identification", "Escherichia coli", "Kosakonia cowanii")
  )
})
test_that("read_biotyper_report works properly with correct dataset and all hits", {
  expect_equal(
    nrow(read_biotyper_report(biotyper, best_hits = FALSE)), 3
  )
  expect_equal(
    ncol(read_biotyper_report(biotyper, best_hits = FALSE)), 52
  )
})
test_that("read_biotyper_report is empty when no peaks are found", {
  expect_equal(
    nrow(read_biotyper_report(biotyper_empty)), 0
  )
  expect_warning(
    read_biotyper_report(biotyper_empty), "Remove"
  )
})
```

## Importing multiple reports

During large-scale analysis, batches of identification are run and can easily be imported using the `read_many_biotyper_reports` function along with their custom-made metadata.

Below is an example of such usage, where one report was artificially extended into multiple reports.

```{r function-read_many_biotyper_reports}
#' Importing a list of Bruker MALDI Biotyper CSV reports
#'
#' @param path_to_reports A vector of paths to the csv files to be imported by [read_biotyper_report()].
#' @param report_ids A vector of character names for each of the reports.
#' @param best_hits A logical indicating whether to return only the best hit in the [read_biotyper_report()] function.
#' @param ... Name-value pairs to be passed on to [dplyr::mutate()]
#'
#' @return A tibble just like the [read_biotyper_report()] function except for an additional column `name` with the `report_ids` used as a prefix of the `spot` name.
#'
#' @seealso [read_biotyper_report()]
#'
#' @note The report identifiers are sanitised to convert all dashes (`-`) as underscores (`_`).
#'
#' @export
#'
#' @examples
read_many_biotyper_reports <- function(path_to_reports, report_ids, best_hits = TRUE, ...) {
  # Import the Bruker Biotyper reports as a named list
  breports <- lapply(
    path_to_reports,
    read_biotyper_report,
    best_hits
  )
  names(breports) <- report_ids
  # Conversion of a named list of dataframe to the dataframe with the name as
  #  a column is now super easy with enframe()
  tibble::enframe(breports) %>%
    tidyr::unnest("value") %>%
    dplyr::mutate(
      "name" = paste(gsub("-", "_", .data$name), .data$spot, sep = "_"),
      "bruker_species" = if_else(
        .data$bruker_species == "not reliable identification",
        paste(.data$bruker_species, .data$name), # make the unmatched species unique for later counts
        .data$bruker_species
      ),
      ...
    ) %>%
    return()
}
```
  
```{r example-read_many_biotyper_reports}
# List of Bruker MALDI Biotyper reports
reports_paths <- system.file(
  c("biotyper.csv", "biotyper.csv", "biotyper.csv"),
  package = "maldipickr"
)
# Read the list of reports and combine them in a single tibble
read_many_biotyper_reports(
  reports_paths,
  report_ids = c("first", "second", "third"),
  growth_temperature = 37.0 # Additional metadata passed to dplyr::mutate
)
```
  
```{r tests-read_many_biotyper_reports}
dummy_multiple_paths <- system.file(
  c("biotyper.csv", "biotyper_empty.csv", "biotyper.csv"),
  package = "maldipickr"
)
test_that("read_many_biotyper_reports works", {
  expect_equal(
    nrow(
      read_many_biotyper_reports(
        dummy_multiple_paths,
        report_ids = c("first", "second", "third")
      )
    ), 6
  )
})
test_that("read_many_biotyper_reports works with arguments passed to mutate", {
  expect_equal(
    read_many_biotyper_reports(
      dummy_multiple_paths,
      report_ids = c("first", "second", "third"),
      growth_temperature = 37.0
    ) %>%
      dplyr::pull(growth_temperature) %>% unique(), 37
  )
})
```

# Importing spectra from the Bruker MALDI Biotyper device

Other than the identification reports, the Bruker MALDI Biotyper device outputs the raw data used for the identification (if not the database) in the form of mass spectra.
Thankfully, the [`MALDIquant`](https://strimmerlab.github.io/software/maldiquant) and [`MALDIquantForeign`](https://strimmerlab.github.io/software/maldiquant) packages helps users import and manipulate these data in R.


## Importing multiple spectra from a directory hierarchy

However, when the Bruker MALDI Biotyper device produces `acqus` files (instead of the native `acqu` files), the `importBrukerFlex()` function from the [`MALDIquantForeign`](https://strimmerlab.github.io/software/maldiquant) package
 will fail with the following error message:

```
Error in .readAcquFile(fidFile = fidFile, verbose = verbose) :
File ‘/data/maldi_dir/targetA/0_D10/1/1SLin/acqu’ doesn't exists!
```

The following `import_biotyper_spectra` used in the example below circumvent this error by creating a symbolic link and conveniently helps removing calibration samples.

The toy dataset bundled with this package is a subset of a dataset in the `MALDIquantExamples` package and consist here of six spectra:
* 1 replicate of species 1
* 2 replicates of species 2
* 3 replicates of species 3

```{r function-import_biotyper_spectra}
#' Importing spectra from the Bruker MALDI Biotyper device
#'
#' This function is a wrapper around the [MALDIquantForeign::importBrukerFlex()] to read both `acqus` and `acqu` MALDI files.
#'
#'
#' @details When using [MALDIquantForeign::importBrukerFlex()] on `acqus` files (instead of the native `acqu` files), the function will fail with the following error message:
#'
#' ```
#' Error in .readAcquFile(fidFile = fidFile, verbose = verbose) :
#' File ‘/data/maldi_dir/targetA/0_D10/1/1SLin/acqu’ doesn't exists!
#' ```
#'
#' But it turns out that `acqu` and `acqus` files [are the same]([https://github.com/sgibb/readBrukerFlexData/wiki/acqu-file), so the function here create `acqu` symbolic links that point to `acqus` files.
#'
#' @param biotyper_directory A path to the folder tree with the spectra to be imported.
#' @param remove_calibration A vector of characters used as regex to indicate which (calibration) spectra are going to be removed.
#'
#' @return A list of [MALDIquant::MassSpectrum] objects
#'
#' @export
import_biotyper_spectra <- function(biotyper_directory, remove_calibration = c("BTS", "Autocalibration")) {
  # List all the acqus files
  acqus_files <- list.files(
    biotyper_directory,
    pattern = "acqus$",
    recursive = TRUE, full.names = TRUE
  )
  # Prepare the absolute filename for acqu
  acqu_files <- gsub("acqus$", "acqu", acqus_files)
  # Acqu files as symbolic links if they do not exist
  do_symbolic_links <- !file.exists(acqu_files)
  if (any(do_symbolic_links)) {
    links_status <- file.symlink(
      from = acqus_files[do_symbolic_links],
      to = acqu_files[do_symbolic_links]
    )
    if (any(!links_status)) {
      stop(
        "Symbolic links could not be created for the files:",
        acqus_files[!links_status]
      )
    }
  }
  biotyper_list <- MALDIquantForeign::importBrukerFlex(biotyper_directory)
  # Catch the arguments of which spectrum to remove
  remove_calibration <- match.arg(remove_calibration, several.ok = T)
  if (length(remove_calibration) != 0) {
    # Regex to remove the chosen spectrum
    to_catch <- c(
      "BTS" = "BTS|BTS_Validation",
      "Autocalibration" = "Autocalibration"
    )
    regex_to_catch <- match.arg(remove_calibration, to_catch, several.ok = T) %>%
      paste(collapse = "|")
    to_remove <- sapply(biotyper_list, function(x) grepl(regex_to_catch, MALDIquant::metaData(x)$file))
    biotyper_list <- biotyper_list[!to_remove]
  }
  return(biotyper_list)
}
```
  
```{r example-import_biotyper_spectra}
# Get an example directory of six Bruker MALDI Biotyper spectra
directory_for_biotyper_spectra <- system.file("toy-species-spectra", package = "maldipickr")
# Import the six spectra
spectra_list <- import_biotyper_spectra(directory_for_biotyper_spectra)
# Display the list of spectra
spectra_list
```
  
```{r tests-import_biotyper_spectra}
directory_for_biotyper_spectra <- system.file("toy-species-spectra", package = "maldipickr")
test_that("import_biotyper_spectra works", {
  expect_length(
    import_biotyper_spectra(directory_for_biotyper_spectra),
    6
  )
})
```

## Evaluate the quality of the spectra

Once the spectra are imported, the `check_spectra` can easily assess whether all the spectra in the list are not empty, of the same length and correspond to profile data.
If some spectra do not satisfy these criteria, the function will exit with a warning and indicate the faulty spectra.
Either way, the function outputs a list of logical vectors (`TRUE` or `FALSE`) indicating whether each of the spectra are empty (`is_empty`), of an odd length (`is_outlier_length`) or not a profile spectra (`is_not_regular`).

```{r function-check_spectra}
#' Evaluate the spectra regularities
#'
#' Assess whether all the spectra in the list are not empty, of the same length and correspond to profile data.
#'
#' @param spectra_list A list of [MALDIquant::MassSpectrum] objects
#'
#' @return A list of logical vectors of length `spectra_list` indicating if the spectra are empty (`is_empty`), of an odd length (`is_outlier_length`) or not a profile spectra (`is_not_regular`).
#'
#' @export
check_spectra <- function(spectra_list) {
  # Checking spectra are not empty
  empty_spectra <- sapply(spectra_list, MALDIquant::isEmpty)
  # Checking spectra are the same length
  ## Getting the most common length
  common_length <- sapply(spectra_list, length) %>%
    # gives length as character/names and occurrence as value
    table() %>%
    which.max() %>%
    # extract length
    names() %>%
    strtoi()
  length_spectra <- sapply(spectra_list, length) == common_length
  # Checking spectra are profile data
  regular_spectra <- sapply(spectra_list, MALDIquant::isRegular)
  # Summarise the checks
  checking_list <- list(
    is_empty = empty_spectra,
    is_outlier_length = !length_spectra,
    is_not_regular = !regular_spectra
  )
  if (any(Reduce(any, checking_list))) {
    # List all the spectra that should be removed
    to_remove <- lapply(checking_list, which) %>%
      unlist() %>%
      unique() %>%
      paste(collapse = ",")
    warning(
      "Some spectra (", to_remove, ") are incorrect (empty, outlier length or irregular).\n",
      "They can be removed as follow:\n",
      " spectra_list <- spectra_list[-c(", to_remove, ")]"
    )
  }
  return(checking_list)
}
```
  
```{r example-check_spectra}
# Get an example directory of six Bruker MALDI Biotyper spectra
directory_for_biotyper_spectra <- system.file("toy-species-spectra", package = "maldipickr")
# Import the six spectra
spectra_list <- import_biotyper_spectra(directory_for_biotyper_spectra)
# Display the list of checks, with FALSE where no anomaly is detected
check_spectra(spectra_list)
# The overall sanity can be checked with Reduce
Reduce(any, check_spectra(spectra_list)) # Should be FALSE
```
  
```{r tests-check_spectra}
test_that("check_spectra works", {
  expect_type(
    check_spectra(
      c(MALDIquant::createMassSpectrum(1:3, 1:3), MALDIquant::createMassSpectrum(11:13, 11:13))
    ), "list"
  )
})
test_that("check_spectra warns with empty spectra", {
  expect_warning(
    check_spectra(c(MALDIquant::createMassSpectrum(mass = 0, intensity = 0))),
    "Some spectra"
  )
})
```
  

```{r development-inflate, eval=FALSE}
# Run but keep eval=FALSE to avoid infinite loop
# Execute in the console directly
fusen::inflate(
  flat_file = "dev/import-data.Rmd",
  vignette_name = c(
    "Import data from Bruker MALDI Biotyper" = "A-import-data-from-bruker-maldi-biotyper"
  )
)
```
